Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2022-12-27 paper-1/manuscript.tex
Rev sec Results:F6

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n = 800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99, p = 0).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
% TODO 5-8: Re-label stim locations once per figure
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike, in response to stimulation at site B in any condition. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (dev - msc) and the contrast in $V_{TA}$ at the start of B trials. Neurons are colored by latency rank, in the sense that the earliest 100 responders are shown as red, and the later neurons as blue. Inset histograms show the distributions and estimated median of the corresponding values across these two categories.
\textbf{c} Mean non-target response contrast (dev - msc) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labelled "late", across networks and stimuli.
\textbf{c} Mean contrast (dev - msc) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli

% TODO rev Fig6 text
Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the more frequent presentation of B changes adaptation, likely reducing the average response to B.
In order to separate these two, we split the non-target contrast shown in Figure \ref{fig5}b into two components: a replacement effect, contrasting the response to B in the msc sequence with the response to all non-target stimuli in the same sequence; and an adaptation effect, contrasting the response to B in dev with the response to B in msc. For the sample network, these contrasts are shown in Figure \ref{fig6}a. Unsurprisingly, replacement yielded a decreased response near the stimulation sites for C, D and E, and an increased response in areas near the stimulation site for B. By contrast, adaptation caused a decreased response in virtually the entire network, particularly in the vicinity of the stimulation site for B.

Across networks and stimuli (Figure \ref{fig6}b), we saw a wide range of replacement effects, including both increased and decreased firing, likely as a result of differing sensitivities of the networks to the various non-target stimuli. On average, stimulus replacement did not have a significant effect (t = -0.255, p = 0.799; two-sided t-test). In contrast, adaptation yielded a reduced response on average (t = -9.01, p = 2.02e-15; one-sided t-test), indicating that the more frequent presentation of B indeed lowered the network's response to this stimulus.

Finally, given the clear spatial pattern exhibited by the replacement effect in the sample network, we surmised that adaptation is driven mainly by neurons close to the stimulation site B responding more frequently due to the more frequent presentation of the stimulus. To show this, we first identified all neurons located within a single excitatory synapse of the stimulated neurons (cf. Figure \ref{fig1}c), and calculated the average replacement effect for these alone. In this subpopulation, replacement yielded a clear response increase, shown in the central box of Figure \ref{fig6}b (t = 8.21, p = 1.5e-13; one-sided t-test). As shown in Figure \ref{fig6}c, this response increase directly translated into elevated thresholds in dev trials compared to msc (t = 5.01, p = 9.59e-07).

Earlier, we showed that the early response is strongly influenced by TA (Figure \ref{fig4}h). Here, we showed that $V_{TA}$ near the stimulation site was increased in dev as a result of the more frequent presentation of B. We posit that this increase in TA affects the early response, reducing the response further downstream and freeing up the rest of the network to respond more strongly to the infrequent target stimulus.

To conclude, TA plays two complementary roles in deviance detection, affecting the standard and deviant response separately: In its first role, as we have shown with the adaptation contrast above, TA reduces the response to frequently presented stimuli regardless of context, leading to a stronger response to infrequently presented stimuli (SSA). Since this happens without the need for large-scale adaptation -- only a small fraction of the network near the stimulation site adapts strongly to frequently presented inputs -- the network as a whole is allowed to relax in the oddball condition. In its second role, TA causes network-wide adaptation in the control condition, where many different stimuli are presented, all of which are infrequently presented and therefore able to evoke large responses. Deviant detection then arises as a relief from this blanket suppression, or as a result of adaptation shifting from the entire network to a small region around the frequently presented standard.
% END TODO

>>>>

Correlating the non-target response and thresholds excitatory neurons (Figure \ref{fig5}c), we found that the reduced non-target response magnitude in dev (t = -7.61, p = 3.5e-12, median = -0.044; one-sided t-test, n = 800) almost perfectly predicted the resulting reduction in threshold adaptation levels (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in Figure \ref{fig5}d, with a median correlation coefficient of 0.994 and a consistent reduction in the average non-target response between msc and dev sequences (t = -11.6, p = 1.46e-21, median = -0.058; one-sided t-test, n = 120), clearly validating our notion that the non-target trial response is the primary driver of threshold adaptation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
% TODO 5-8: Re-label stim locations once per figure
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any condition. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (dev - msc) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are plotted with crosses, while later responders are plotted dots. Inset histograms show the distributions of the corresponding values across these two categories, and the confidence intervals of their medians, calculated by bootstrap with 10000 iterations.
\textbf{c} Median non-target response contrast (dev - msc) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labelled "late", across networks and stimuli.
\textbf{c} Median contrast (dev - msc) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli

Finally, we examined the cause of the reduced response in non-target trials. Going from the msc sequence to the dev sequence, two things change: Firstly, stimulations with non-target C, D and E are replaced with B, which would cause different responses even in the absence of adaptation. Secondly, the resulting  frequent presentation of B changes TA values, likely causing adaptation to B and therefore decreasing the average response. We hypothesized that this adaptation is mediated by a small subset of neurons that (a) responds very quickly to stimulation in B, (b) responds more strongly to non-target stimuli in dev than msc, and (c) has increased TA levels as a result, reducing the later response in the same way as demonstrated in Figure \ref{fig4} in reverse.

To see this in the sample network, consider first the latency of responses to B across all conditions, shown by way of rank order in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward from the stimulus site. We did notice that the early response appears to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B in any condition, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the response contrast (dev - msc) to the contrast in $V_{TA}$ at the start of B trials. Across all neurons, a strong linear relationship (Pearson's $\rho$ = 0.98) echoed the related finding in Figure \ref{fig5}c. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the dev sequence, while the network as a whole decreased. Correspondingly, the TA levels of this portion of the network also tended to increase, in clear contrast to the network as a whole.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that the non-target response and $V_{TA}$ contrasts were robustly correlated (Median Pearson's $\rho$ 0.996, T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon test for $\rho > 0$, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast (Figure \ref{fig6}c) and the median contrast in $V_{TA}$ at the start of B trials (Figure \ref{fig6}d) within these two portions. We found that the response in the early portion was consistently greater in dev than msc (T = 6.98e+03, p = 8.48e-19; one-sided Wilcoxon, n = 120), unlike the late portion, which decreased (T = 295, p = 1.23e-18; see also Figure \ref{fig5}d). The contrast in $V_{TA}$ followed the same pattern, with the early portion exhibiting stronger TA in dev (T = 4.63e+03, p = 2.21e-17), while the late portion was more recovered (T = 26, p = 2.97e-17).

To sum up, we showed in Figure \ref{fig4} that greater $V_{TA}$ in the portion of the network responding earliest to a stimulus plays an important role in reducing the overall response. Here, we showed that the neurons responding earliest to B exhibit higher $V_{TA}$ in the dev sequence as a consequence of an increased response, in direct contrast to the response and TA levels of the network as a whole.

Thus, we have finally traced the greater response to A in the dev sequence to its roots, and revealed two complementary roles of TA: The frequent presentation of non-target stimulus B in the dev sequence and concomitant increased activity in a subset of neurons responding early to B causes \emph{local adaptation} by TA in this subset, leading to smaller network responses. By contrast, stimuli in the msc sequence are presented more sporadically, allowing local adaptation to decay, thus leading to larger network responses. This difference in non-target response sizes -- small in dev, large in msc -- is then reflected in network-wide \emph{global adaptation}, such that the network is more excitable to stimulation with A (or other non-standard inputs) in the dev sequence.
